---
title: "HW3: Predictors of Ever Being Married"
subtitle: "Household Pulse Survey"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(broom)

load("Household_Pulse_data.RData")
d <- Household_Pulse_data
```

## Building the outcome variable

The survey's `MS` variable records current marital status: married, widowed,
divorced, separated, or never married. "Ever married" groups together anyone
who is or has been married (married / widowed / divorced / separated) versus
those who report never having married. Rows with missing `MS` are dropped.

```{r}
d <- d %>%
  filter(!is.na(MS)) %>%
  mutate(
    ever_married = if_else(MS == "never", "Never married", "Ever married"),
    ever_married = factor(ever_married, levels = c("Never married", "Ever married")),

    # Household Pulse Survey covers 2020-2024; using the midpoint year as a
    # reference since this extract has no per-row survey date.
    age = 2023 - TBIRTH_YEAR,
    age_group = cut(age, breaks = c(17, 24, 34, 44, 54, 64, 74, 100),
                     labels = c("18-24", "25-34", "35-44", "45-54",
                                "55-64", "65-74", "75+")),

    # Kept as regular (non-ordered) factors so the regression below reports
    # interpretable dummy-variable coefficients against a reference level,
    # rather than orthogonal polynomial contrasts.
    education = factor(EEDUC, levels = c("less than hs", "some hs", "HS diploma",
                                          "some coll", "assoc deg", "bach deg", "adv deg")),

    income = factor(INCOME, levels = c("HH income less than $25k", "HH income $25k - $34.9k",
                                        "HH income $35k - 49.9", "HH income $50k - 74.9",
                                        "HH income $75 - 99.9", "HH income $100k - 149",
                                        "HH income $150 - 199", "HH income $200k +")),

    gender = factor(EGENID_BIRTH),
    race = factor(RRACE),
    hispanic = factor(RHISPANIC),
    region = factor(REGION),

    # This extract encodes "not applicable" as a literal "NA" factor level
    # rather than a true missing value, so those levels must be dropped
    # explicitly rather than relying on is.na().
    employment = factor(ANYWORK, levels = c("yes employment in last 7 days",
                                             "no employment in last 7 days")),

    has_kids = if_else(as.character(KIDS_LT5Y) != "NA" |
                          as.character(KIDS_5_11Y) != "NA" |
                          as.character(KIDS_12_17Y) != "NA",
                        "Has children in HH", "No children in HH"),
    has_kids = factor(has_kids)
  ) %>%
  filter(age >= 18, age <= 95)

n_total <- nrow(d)
overall_rate <- mean(d$ever_married == "Ever married")
```

Sample size after cleaning: **`r comma(n_total)`** respondents. Overall,
**`r percent(overall_rate, accuracy = 0.1)`** report having ever been married.

## Ever-married rate by demographic factor

A small helper summarizes the ever-married rate within levels of a grouping
variable, then plots it as a bar chart.

```{r}
plot_rate_by <- function(data, group_var, title, xlab) {
  summary_tbl <- data %>%
    group_by(.group = {{ group_var }}) %>%
    summarise(rate = mean(ever_married == "Ever married"), n = n(), .groups = "drop") %>%
    filter(!is.na(.group))

  ggplot(summary_tbl, aes(x = .group, y = rate)) +
    geom_col(fill = "#2C5F8A") +
    geom_text(aes(label = percent(rate, accuracy = 1)), vjust = -0.4, size = 3.2) +
    scale_y_continuous(labels = percent, limits = c(0, 1)) +
    labs(title = title, x = xlab, y = "Share ever married") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Age

```{r}
p_age <- plot_rate_by(d, age_group, "Ever Married by Age Group", "Age group")
p_age
ggsave("hw3_married_by_age.png", p_age, width = 8, height = 5, dpi = 300)
```

### Education

```{r}
p_educ <- plot_rate_by(d, education, "Ever Married by Education", "Education")
p_educ
ggsave("hw3_married_by_education.png", p_educ, width = 8, height = 5, dpi = 300)
```

### Household income

```{r}
p_income <- plot_rate_by(d, income, "Ever Married by Household Income", "Household income")
p_income
ggsave("hw3_married_by_income.png", p_income, width = 9, height = 5, dpi = 300)
```

### Gender, race/ethnicity, and region

```{r}
p_gender <- plot_rate_by(d, gender, "Ever Married by Gender", "Gender")
p_gender

p_race <- plot_rate_by(d, race, "Ever Married by Race", "Race")
p_race

p_hispanic <- plot_rate_by(d, hispanic, "Ever Married by Hispanic Origin", "Hispanic origin")
p_hispanic

p_region <- plot_rate_by(d, region, "Ever Married by Region", "Region")
p_region

ggsave("hw3_married_by_gender.png", p_gender, width = 6, height = 5, dpi = 300)
ggsave("hw3_married_by_race.png", p_race, width = 7, height = 5, dpi = 300)
ggsave("hw3_married_by_hispanic.png", p_hispanic, width = 6, height = 5, dpi = 300)
ggsave("hw3_married_by_region.png", p_region, width = 7, height = 5, dpi = 300)
```

### Children in household and employment

```{r}
p_kids <- plot_rate_by(d, has_kids, "Ever Married by Presence of Children", "Children in household")
p_kids

p_emp <- plot_rate_by(d, employment, "Ever Married by Employment Status", "Employment (last 7 days)")
p_emp

ggsave("hw3_married_by_kids.png", p_kids, width = 6, height = 5, dpi = 300)
ggsave("hw3_married_by_employment.png", p_emp, width = 7, height = 5, dpi = 300)
```

## Which factors predict ever having been married?

The bar charts above show each factor on its own, but many of these variables
are correlated with each other (age and income, for instance). A logistic
regression lets us see which factors still predict marital history once we
account for the others.

```{r}
model <- glm(
  ever_married ~ age + gender + race + hispanic + education + income +
    region + has_kids + employment,
  data = d,
  family = binomial
)

summary(model)
```

```{r}
or_tbl <- tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

or_tbl
```

```{r}
p_or <- ggplot(or_tbl, aes(x = estimate, y = fct_reorder(term, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "#2C5F8A") +
  scale_x_log10() +
  labs(title = "Predictors of Ever Being Married (Logistic Regression)",
       subtitle = "Odds ratios with 95% confidence intervals, log scale",
       x = "Odds ratio", y = NULL) +
  theme_minimal(base_size = 12)

p_or
ggsave("hw3_odds_ratios.png", p_or, width = 9, height = 7, dpi = 300)
```

## Takeaways

- Report the overall ever-married rate and how it varies across the strongest
  bivariate factors from the bar charts (age is typically the biggest driver,
  since marital status accumulates over the life course).
- Use the odds-ratio plot to say which factors remain significant predictors
  after controlling for the others — points whose confidence interval crosses
  1 are not statistically distinguishable from "no effect" in this model.
- Note the age proxy caveat: this extract has no per-row survey year, so age
  is approximated using the 2020-2024 survey midpoint (2023).
